TBD

### 2018-10-18 paper/paper.tex
more math

\section{gaussian processes and the celerite model}
	A Gaussian process is a stochastic noise model which consists of a mean:
	\begin{equation} 
		\bvec{\mu_\theta}(\bvec{x}) = \expandvec{\mu_\bvec{\theta}(x_1)}{\mu_\bvec{\theta}(x_N)}
	\end{equation}
	and covariance defined the kernel function $k_\bvec{\alpha}(x_n, x_m)$. The mean and kernel functions depend on the parameters 
	$\bvec{\theta}$ and $\bvec{\alpha}$ respectively, which are referred to as ``hyperparameters" of the GP. The kernel function defines a covariance matrix 
	$\left[K\right]_{n, m} = k_\alpha(x_n, x_m)$. This model has a likelihood function given by
	\begin{equation}
		\ln\ \mathcal{L(\bvec{\theta}, \bvec{\alpha})} = \ln\ p(\bvec{y}|X, \bvec{\theta}, \bvec{\alpha}) = 
			-\frac{1}{2}\bvec{r_\theta}^\T K_\bvec{\alpha}^{-1}\bvec{r_\theta} 
			-\frac{1}{2}\ln\ \mathrm{det}(K_\bvec{\alpha}) - \frac{N}{2}\ln(2\pi)
	\end{equation}
	\gordontodo{I'm using $X$ here because DFM does, but wouldn't $\bvec{x}$ be more clear?}

	where 
	\begin{equation}
		\bvec{y} = \expandvec{y_1}{y_N}
	\end{equation}
	are the data, $N$ is the number of datapoints, and
	\begin{equation}
		\bvec{r_\theta} = \bvec{y}-\bvec{\mu_\theta}(\bvec{x})
	\end{equation}
	By maximizing the likelihood (or, in practice, minimizing the negative of the log-likelihood), one can obtain an estimate of the hyperparameters $\bvec{\theta}$ and 
	$\bvec{\alpha}$. Furthermore, Markov Chain Monte Carlo (MCMC) methods can be used to sample the likelihood in hyperparameter-space. \gordontodo{expand 
	on this.}
	\subsection{The \celerite model}
		The primary limitation of Gaussian process noise models is that they are computationally expensive, with computation of the likelihood function scaling as 
		$\mathcal{O}(N^3)$ due to the matrix inversion. This is especially problematic for use cases such as MCMC which require calling the likelihood function many 
		times. This has prevented the adoption of Gaussian process noise models for datasets larger than $N \approx 10^3$, except in cases where approximate 
		methods suffice. The \celerite method overcomes this issue by reducing the time to compute the likelihood function to $\mathcal{O}(NJ^2)$ for one 
		class of likelihood functions; where J is the number of \celerite terms making up the kernel function of the GP. This class of kernel functions are given by 
		\begin{equation}
			k_\alpha(x_n, x_m) = \sigma_n^2 \delta_{nm} + \sum_{j=1}^J a_j e^{-c_j\tau_{nm}}
		\end{equation}
		where $\bvec{\alpha} = (a_1...a_j, c_1...c_j)$, $\sigma_n^2$ is the variance of the white noise, and $\tau_{nm} = |x_n-x_m|$. 
		The coefficients $a$ and $c$ may be complex.

@@

\section{2d gaussian processes}
	\subsection{kronecker structure of the 2d covariance matrix}
	In two dimensions the kernel function depends on two coordinates. Whereas above we used $\bvec{x}$ to refer to the single coordinate, we will now use
	\begin{equation}
		\bvec{t} = \expandvec{t_1}{t_N}
	\end{equation}
	to refer to the first coordinate, and
	\begin{equation}
		\bvec{u} = \expandvec{u_1}{u_M}
	\end{equation}
	to refer to the second coordinate. In most cases, we will demand that $M << N$. We then redefine $\bvec{x}$ to be the list of points
	\begin{equation}
		\bvec{x} = \expandvec{(t_1, u_1), (t_1, u_2), \cdots (t_1, u_M)}{(t_N, u_M)}
	\end{equation}
	We can also write this as
	\begin{equation}
		x_i = (t_{\lfloor(i-1)/M\rfloor + 1}, u_{i - \lfloor(i-1)/M\rfloor M})
	\end{equation} 
	The covariance between $x_i$ and $x_j$ is then given by
	\begin{equation}
		k_{\bvec{\alpha}\bvec{\beta}}(x_i, x_j) = \sigma_i^2\delta_{ij} + q_\bvec{\alpha}(\tau_{ij})\circ r_\bvec{\beta}(\nu_{ij})
	\end{equation}
	where $\tau_{ij} = |t_{\lfloor(i-1)/M\rfloor + 1}-t_{\lfloor(j-1)/M\rfloor + 1}|$, 
	$\nu_{ij} = |u_{i - \lfloor(i-1)/M\rfloor M}-u_{j - \lfloor(j-1)/M\rfloor M}|$; 
	$q$ and $r$ are one-dimensional kernel functions which depend on the hyperparameters $\bvec{\alpha}$ and 
	$\bvec{\beta}$ respectively; 
	and $\circ$ is some binary operation. 
	If $Q_\bvec{\alpha}$ and $R_\bvec{\alpha}$ are the covariance matrices corresponding to 
	$q_\bvec{\alpha}$ and $r_\bvec{\alpha}$ such that
	\begin{equation}
		\left[Q_\bvec{\alpha}\right]_{ij} = q_\bvec{\alpha}(\tau_{ij})\ \ \ \mathrm{and}\ \ \ \left[R_\bvec{\beta}\right]_{ij} = r_\bvec{\beta}(\tau_{ij})
	\end{equation} 
	then, using $q_{ij}$ and $r_{ij}$ as shorthand for the entries in $Q_\bvec{\alpha}$ and $R_\bvec{\beta}$, $K_\bvec{\alpha\beta}$ can be represented as: 
	\begin{equation}
		K_\bvec{\alpha\beta} = \Sigma + 
			\begin{bmatrix}
				r_{11}\circ Q_\bvec{\alpha} & ... & r_{1N}\circ Q_\bvec{\alpha} \\
				\vdots & \ddots & \\
				r_{N1}\circ Q_\bvec{\alpha} & ... & r_{NN}\circ Q_\bvec{\alpha}
			\end{bmatrix}
	\end{equation}
	where $[\Sigma]_{ij} = \sigma_i^2\delta_{ij}$ and
	\begin{equation}
		r_{pq}\circ Q_\bvec{\alpha} = 
			\begin{bmatrix}
				r_{pq}\circ q_{11} & ... & r_{pq}\circ q_{1M} \\
				\vdots & \ddots & \\
				r_{pq}\circ q_{M1} & ... & r_{pq}\circ q_{MM}
			\end{bmatrix}
	\end{equation}
	If the operator $\circ$ represents multiplication, we have 
	\begin{equation}
		K_\bvec{\alpha\beta} = \Sigma + R_\bvec{\beta} \otimes  Q_\bvec{\alpha}
	\end{equation}
	where $\otimes$ is the Kronecker product. 
	\subsection{performance}

>>>>

\section{gaussian processes and the celerite model}
	A Gaussian process is a stochastic noise model which consists of a mean:
	\begin{equation} 
		\bvec{\mu_\theta}(\bvec{x}) = \expandvec{\mu_\bvec{\theta}(x_1)}{\mu_\bvec{\theta}(x_N)}
	\end{equation}
	and covariance defined the kernel function $k_\bvec{\alpha}(x_n, x_m)$. The mean and kernel functions depend on the parameters 
	$\bvec{\theta}$ and $\bvec{\alpha}$ respectively, which are referred to as ``hyperparameters" of the GP. The kernel function defines a covariance matrix 
	$\left[K\right]_{n, m} = k_\alpha(x_n, x_m)$. This model has a likelihood function given by
	\begin{equation}
		\ln\ \mathcal{L(\bvec{\theta}, \bvec{\alpha})} = \ln\ p(\bvec{y}|X, \bvec{\theta}, \bvec{\alpha}) = 
			-\frac{1}{2}\bvec{r_\theta}^\T K_\bvec{\alpha}^{-1}\bvec{r_\theta} 
			-\frac{1}{2}\ln\ \mathrm{det}(K_\bvec{\alpha}) - \frac{N}{2}\ln(2\pi)
	\end{equation}
	\gordontodo{I'm using $X$ here because DFM does, but wouldn't $\bvec{x}$ be more clear?}
	where 
	\begin{equation}
		\bvec{y} = \expandvec{y_1}{y_N}
	\end{equation}
	are the data, $N$ is the number of datapoints, and
	\begin{equation}
		\bvec{r_\theta} = \bvec{y}-\bvec{\mu_\theta}(\bvec{x})
	\end{equation}
	By maximizing the likelihood (or, in practice, minimizing the negative of the log-likelihood), one can obtain an estimate of the hyperparameters $\bvec{\theta}$ and 
	$\bvec{\alpha}$. Furthermore, Markov Chain Monte Carlo (MCMC) methods can be used to sample the likelihood in hyperparameter-space. \gordontodo{expand 
	on this.}
	\subsection{The \celerite model}
		The primary limitation of Gaussian process noise models is that they are computationally expensive, with computation of the likelihood function scaling as 
		$\mathcal{O}(N^3)$ due to the matrix inversion. This is especially problematic for use cases such as MCMC which require calling the likelihood function many 
		times. This has prevented the adoption of Gaussian process noise models for datasets larger than $N \approx 10^3$, except in cases where approximate 
		methods suffice. The \celerite method overcomes this issue by reducing the time to compute the likelihood function to $\mathcal{O}(NJ^2)$ for one 
		class of kernel functions; where J is the number of \celerite terms making up the kernel function of the GP. This class of kernel functions are given by 
		\begin{equation}
			k_\alpha(t_n, t_m) = \sigma_n^2 \delta_{nm} + \sum_{j=1}^J a_j e^{-c_j\tau_{nm}}
		\end{equation}
		where $\bvec{\alpha} = (a_1...a_j, c_1...c_j)$, $\sigma_n^2$ is the variance of the white noise, and $\tau_{nm} = |t_n-t_m|$. 
		The coefficients $a$ and $c$ may be complex.

		For kernel functions of this form, the covariance matrix can be written as a semi-separable matrix of rank $2J$:
		\begin{equation}
			K = A + \mathrm{tril}(UV^\T) + \mathrm{triu}(VU^\T)
		\end{equation}
		where $U$ and $V$ have entries given by: 
		\begin{eqnarray}
			U_{n, 2j-1} &=& a_je^{-c_jt_n}\cos(d_jt_n) + b_je^{-c_jt_n}\sin(d_jt_n) \nonumber \\
			U_{n, 2j} &=& a_je^{-c_jt_n}\sin(d_jt_n) - b_je^{-c_jt_n}\cos(d_jt_n) \nonumber \\
			V_{m, 2j-1} &=& e^{c_jt_m}\cos(d_jt_m) \nonumber \\
			V_{m, 2j} &=& e^{c_jt_m}\sin(d_jt_m)
		\end{eqnarray}
		and A is a diagonal matrix:
		\begin{equation}
			A_{n,n} = \sigma_n^2 + \sum_{j=1}^Ja_j
		\end{equation}
		Cholesky decomposition can be performed on this matrix in $\mathcal{O}(NJ^2)$ time, as described in the appendix.

@@

\section{2d gaussian processes}
	\subsection{kronecker structure of the 2d covariance matrix}
	In two dimensions the kernel function depends on two coordinates. We will continue to use
	\begin{equation}
		\bvec{t} = \expandvec{t_1}{t_N}
	\end{equation}
	to refer to the first coordinate, and introduce
	\begin{equation}
		\bvec{u} = \expandvec{u_1}{u_M}
	\end{equation}
	to refer to the second coordinate. In most cases, we will demand that $M << N$. We then define $\bvec{x}$ to be the list of points
	\begin{equation}
		\bvec{x} = \expandvec{(t_1, u_1), (t_1, u_2), \cdots (t_1, u_M)}{(t_N, u_M)}
	\end{equation}
	We can also write this as
	\begin{equation}
		x_i = (t_{\lfloor(i-1)/M\rfloor + 1}, u_{i - \lfloor(i-1)/M\rfloor M})
	\end{equation} 
	The covariance between $x_i$ and $x_j$ is then given by
	\begin{equation}
		k_{\bvec{\alpha}\bvec{\beta}}(x_i, x_j) = \sigma_i^2\delta_{ij} + q_\bvec{\alpha}(\tau_{ij})\circ r_\bvec{\beta}(\nu_{ij})
	\end{equation}
	where $\tau_{ij} = |t_{\lfloor(i-1)/M\rfloor + 1}-t_{\lfloor(j-1)/M\rfloor + 1}|$, 
	$\nu_{ij} = |u_{i - \lfloor(i-1)/M\rfloor M}-u_{j - \lfloor(j-1)/M\rfloor M}|$; 
	$q$ and $r$ are one-dimensional kernel functions which depend on the hyperparameters $\bvec{\alpha}$ and 
	$\bvec{\beta}$ respectively; 
	and $\circ$ is some binary operation. 
	If $Q_\bvec{\alpha}$ and $R_\bvec{\alpha}$ are the covariance matrices corresponding to 
	$q_\bvec{\alpha}$ and $r_\bvec{\alpha}$ such that
	\begin{equation}
		\left[Q_\bvec{\alpha}\right]_{ij} = q_\bvec{\alpha}(\tau_{ij})\ \ \ \mathrm{and}\ \ \ \left[R_\bvec{\beta}\right]_{ij} = r_\bvec{\beta}(\tau_{ij})
	\end{equation} 
	then, using $q_{ij}$ and $r_{ij}$ as shorthand for the entries in $Q_\bvec{\alpha}$ and $R_\bvec{\beta}$, $K_\bvec{\alpha\beta}$ can be represented as: 
	\begin{equation}
		K_\bvec{\alpha\beta} = \Sigma + 
			\begin{bmatrix}
				q_{11}\circ R_\bvec{\beta} & ... & q_{1N}\circ R_\bvec{\beta} \\
				\vdots & \ddots & \\
				q_{N1}\circ R_\bvec{\beta} & ... & q_{NN}\circ R_\bvec{\beta}
			\end{bmatrix}
	\end{equation}
	where $[\Sigma]_{ij} = \sigma_i^2\delta_{ij}$ and
	\begin{equation}
		q_{pq}\circ R_\bvec{\alpha} = 
			\begin{bmatrix}
				q_{pq}\circ r_{11} & ... & q_{pq}\circ r_{1M} \\
				\vdots & \ddots & \\
				q_{pq}\circ r_{M1} & ... & q_{pq}\circ r_{MM}
			\end{bmatrix}
	\end{equation}
	If the operator $\circ$ represents multiplication, we have 
	\begin{equation}
		K_\bvec{\alpha\beta} = \Sigma + Q_\bvec{\alpha} \otimes  R_\bvec{\beta}
	\end{equation}
	where $\otimes$ is the Kronecker product. If the covariance in $\bvec{t}$ is described by a \celerite kernel can write 
	\begin{eqnarray}
		K &=& \Sigma + \left[\mathrm{tril}(UV^\T) + \mathrm{triu}(VU^\T)\right]\otimes R \\
		&=& \Sigma + \mathrm{tril}\left[ (U\otimes R)(V\otimes I_M)^\T\right] + \mathrm{triu}\left[ (V\otimes I_M)(U\otimes R)^\T\right]
	\end{eqnarray}
	This covariance has the same semi-separable form as in the one-dimensional case, but now $K$ is an $(NM\times NM)$ matrix of rank $2JM$. 
	This means that we can apply the same Cholesky decomposition algorithm (described in the appendix) as in the 1d case to compute the GP in 
	$\mathcal{O}(NJ^2M^3)$.

	\subsection{performance}
	Computation of a \celerite GP takes $\mathcal{O}(NJ)$ where $N$ is the length of the coordinate vector $\bvec{x}$ and $J$ is the number of 
	\celerite terms making up the kernel function.
